\subsection{Tangents and Areas} 

Building on his concept of \textit{fluxions}, Newton sought to explain the celestial mechanics that Kepler had described — particularly the Second Law: that a planet sweeps out equal areas in equal intervals of time.

But for Newton, this was more than an empirical fact — it was a geometric key.

Newton imagined a planet tracing a polygonal path around the sun, changing direction at small but finite intervals. The motion between each point was along a straight line — the tangent to the curve at that instant. A central force, acting in discrete impulses, would “bend” the path at each vertex toward the sun.

This construction allowed him to prove that if a body is always deflected toward a central point (like the Sun), then the areas swept out by its radius vector are equal over equal times.

Newton took Kepler’s geometric observation and pushed it into the realm of the infinitesimal. Instead of finite triangles, he considered the area swept out over an infinitesimal time \( \Delta t \), and then took the limit as \( \Delta t \to 0 \):

\[
\lim_{\Delta t \to 0} \frac{\Delta A}{\Delta t} = \frac{1}{2} r^2 \frac{d\theta}{dt}
\]

This quantity — the instantaneous rate of area swept — is not just a computational convenience. It’s a profound physical invariant. No matter where the planet is in its orbit—speeding up near perihelion, slowing down near aphelion—the area it sweeps out per unit of absolute time remains perfectly constant.

That constancy has teeth.

It tells us that the only possible force governing this motion must act \textit{along the radius vector}, directly between the planet and the Sun. Why? Because only a force that pulls straight inward preserves the balance of motion in just the right way to keep the area rate constant. If the force had any sideways component, it would distort the triangle—stretching or squishing the area—violating Kepler’s elegant rule.

So what Kepler described as a poetic geometry, Newton reinterpreted as a diagnostic.
\textbf{The conservation of area isn’t just a pretty fact. It’s evidence.}

It reveals that the invisible hand guiding the orbit must be a \textit{central force}—a force that always aims toward the Sun, regardless of the planet’s speed or position.

In Newton’s hands, Kepler’s Second Law transformed from a mysterious observational regularity into a tool of deduction. It became a geometric fingerprint of force direction—a testable consequence of deeper mechanics.
This was the turning point:
\textbf{Kepler gave us the pattern. Newton found the cause.}

In modern terms, we’d say that angular momentum is conserved. But in Newton’s original fluxional geometry, it was the area rate—the sweep of triangles over time—that carried the truth.
And that truth pointed like an arrow straight to the Sun.

\begin{figure}[H]
\centering
\begin{tikzpicture}[scale=2.5]

  % Sun at origin
  \filldraw[yellow!80!orange] (0,0) circle (0.04) node[below left=2pt] {\textbf{Sun}};

  % Orbit path (approx polygonal)
  \draw[thick, blue] (0:1.2) -- (30:1.1) -- (60:1.25) -- (90:1.1) -- (120:1.2);

  % Radius vectors
  \foreach \a in {0,30,60,90,120} {
    \draw[gray, dashed] (0,0) -- (\a:{1 + 0.2*sin(\a)});
  }

  % Tangents
  \draw[dashed, red] (0:1.2) -- ++(30:0.4);
  \draw[dashed, red] (30:1.1) -- ++(30:0.4);
  \draw[dashed, red] (60:1.25) -- ++(30:0.4);

  % Area sectors (for visual)
  \foreach \a/\b in {0/30, 30/60, 60/90} {
    \fill[blue!20, opacity=0.5] (0,0) -- (\a:{1 + 0.2*sin(\a)}) -- (\b:{1 + 0.2*sin(\b)}) -- cycle;
  }

  \node at (1.2, 0.9) {\scriptsize Equal Area in Equal Time};

\end{tikzpicture}
\caption{Newton’s polygonal construction: a planet moving in equal time steps traces out equal areas. Each segment approximates a tangent; each deflection represents a central force impulse. In the limit of infinitely small time steps, the motion becomes smooth and the area swept per unit time becomes \( \frac{1}{2} r^2 \frac{d\theta}{dt} \), a constant.}
\end{figure}

\subsubsection{From Area to Force} 

In Proposition I of the \textit{Principia}, Newton proved:  

\textit{“A body under a centripetal force sweeps out equal areas in equal times.”}

The converse is also true: if equal areas are swept in equal time, then the force must point toward the center. The notion of “force” is geometric. It’s the repeated deflection of the tangent line by an inward-pulling agent.

\subsubsection{From Fluxions to Gravity.} 

Once Newton had this geometric foundation, he began to explore the nature of the force itself.

In Proposition XI of the \textit{Principia}, he showed that if a planet’s orbit is an ellipse and the force is centripetal, then the force must be inversely proportional to the square of the distance:

\[
F \propto \frac{1}{r^2}
\]

To derive this, Newton analyzed how the velocity (the fluxion of position) changes along the curve. Using geometry and proportions, he established that the acceleration required to maintain the elliptical orbit — which he related to curvature and tangents — must obey this inverse-square law.

This, combined with Kepler’s laws, gave the world its first universal law of gravitation.

\subsubsection{Tangents as the Language of Force.} 

In Newton’s hands, the tangent was more than a line — it was a law. The fluxion $\dot{x}$ described the direction and speed of a moving point; the change of the tangent described acceleration; and acceleration, in Newton’s system, was nothing other than force.

\[
\text{Fluxion} \rightarrow \text{Tangent} \rightarrow \text{Change of Tangent} \rightarrow \text{Force}
\]

This wasn’t just a mathematical trick — it was a physical revolution.


\begin{tcolorbox}[colback=gray!5!white, colframe=black, title=\textbf{Historical Sidebar: Newton’s Fluxions and the Geometry of Instantaneous Motion}, fonttitle=\bfseries, arc=1.5mm, boxrule=0.4pt]

  When Isaac Newton analyzed planetary motion, he built directly on \textbf{Kepler’s Second Law}—that a planet sweeps out \textit{equal areas in equal times}. But where Kepler reasoned in finite steps, Newton introduced the mathematics of the \textit{instant}.
  
  Using his method of \textbf{fluxions}, Newton reinterpreted Kepler’s geometric insight in terms of \textit{momentary motion}. He considered an infinitely small duration of time and the triangular area swept out in that instant.
  
  \vspace{0.5em}
  If a planet is at distance \( r \) from the center and moving with angular fluxion \( \dot{\theta} \), then the rate at which area is swept out—the \textit{fluxion of area}—is:
  
  \[
  \dot{A} = \frac{1}{2} r^2 \dot{\theta}
  \]
  
  This quantity, \( \dot{A} \), remains constant under a centripetal force. Newton took this as more than a computational result—it was a geometric truth in motion.
  
  \vspace{0.5em}
  From this, he concluded that the governing force must be \textbf{directed toward the center}, for only such a force maintains the constancy of \( \dot{A} \).
  
  \[
  \text{Constancy of } \dot{A} \quad \Rightarrow \quad \text{Force directed along the radius}
  \]
  
  Newton thus extracted a law of force from a law of area—not through algebra, but through the geometry of instantaneous change. This was the turning point where discrete planetary geometry became continuous mechanics.
  
\end{tcolorbox}
 
\begin{tcolorbox}[colback=gray!5!white, colframe=black, title=\textbf{Historical Footnote: Philoponus and the Seeds of Inertia}, fonttitle=\bfseries, arc=1.5mm, boxrule=0.4pt]

  Long before Newton formalized motion through fluxions and gravitational forces, \textbf{John Philoponus} (c.~490--570 AD) had already challenged a core idea of Aristotelian physics: the belief that continuous motion required continuous external force.
  
  Philoponus proposed instead the theory of \textbf{impetus}---that a moving object carries an internalized force imparted by its mover. Once set in motion, the object would continue moving without further external influence until that impetus was exhausted, for instance by friction or resistance.
  
  \medskip
  
  This idea---radical in his time---foreshadowed the modern principle of \textbf{inertia}. Where Aristotle saw motion as dependent on an ever-present pusher, Philoponus glimpsed the possibility that motion could be self-sustaining.
  
  \medskip
  
  Although largely ignored for centuries, Philoponus's conceptual shift laid essential philosophical groundwork for later developments in mechanics. When Newton formulated his First Law---that a body in motion remains in motion unless acted upon by an external force---he was, in effect, realizing a vision of motion that Philoponus had first set in motion more than a thousand years earlier.
  
\end{tcolorbox}