\subsection{Observability: Because the Model Was Dying, but Only on Tuesdays}

\textbf{Act I: The Illusion of Stability}

For months, the fraud detection model worked flawlessly.\\
No alerts. No errors. Just vibes.

Until someone noticed a pattern.

Every Tuesday at exactly 3PM, the model started flagging 70\% of transactions as ``high risk.''\\
By 5PM, customer support was overwhelmed.\\
By Wednesday morning? Everything was mysteriously fine again.

\vspace{1em}
\textbf{Act II: The Void}

The team dove in.

The logs? Useless.\\
The metrics? Sparse.\\
The dashboard? Didn't exist.

It was as if the system had gaslit itself into thinking nothing was wrong.

\vspace{1em}
\textbf{Act III: The Curse Revealed}

After weeks of shoulder-shrugging and Slack speculation, someone finally found the culprit:

A weekly cron job, scheduled precisely for Tuesday at 3PM, was reprocessing a batch of transaction metadata.\\
Except it had a timezone bug.

The timestamps were shifted into the future.\\
The model thought every transaction was ``early.'' Suspiciously early.

Not fraud. Just daylight savings.

\vspace{1em}
\textbf{Act IV: The Cult of Debugging}

Without observability, debugging became a ritualistic practice:
\begin{itemize}
    \item Deploy a fix and hope
    \item Read the logs like tea leaves
    \item Pray to the CI gods
\end{itemize}

Root cause analysis devolved into storytelling.\\
Truth became a matter of consensus.

\vspace{1em}
\textbf{Epilogue: Why Observability Matters}

Without proper logging, metrics, and traces, debugging ML systems becomes an occult science.

You end up relying on gut instinct, Slack rumors, and cargo cult rituals.

Observability should answer: What changed? When? Why? \\
And sometimes: Why is the model only racist on weekends?

Don't guess. Instrument.

