\subsection{Ratio and Proportion: The First Math Flex}

Somewhere along the way, people realized magnitudes weren’t just for measuring beer. They could also describe relationships between things. \textbf{Enter ratios}.

A ratio is just a way of saying, This thing is bigger than that thing by some factor. For example:

\begin{quote}
\[
\text{Length} : \text{Width} = 2:1
\]
\end{quote} 


This was useful because it allowed people to build things with some level of predictability -- things like fields, granaries, and temples dedicated to gods who mostly ignored them.  

In ancient Egypt, mathematics was a government job. You can thank the Nile River for that. Every year, the river flooded, wiping out property lines like a chaotic, unregulated HOA.  

This led to the invention of surveyors, also known as the \textit{harpedonaptae} (literally “rope stretchers”), who used ratios to put the land back together.  

The harpedonaptae didn’t have laser levels or GPS satellites. What they had were ropes, stakes, and math; and somehow, they made it work. Their ropes were knotted at regular intervals, which allowed them to create proportional lengths and consistent angles. It was like ancient trigonometry, but without the sines, cosines, or existential dread of taking a calc final.

\begin{figure}[H]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[scale=1, every node/.style={font=\small}]

% --- 3-4-5 Triangle (Left) ---
\draw[thick] (0,0) -- (4,0) -- (4,3) -- cycle;
\draw[fill=gray!10] (4,0) rectangle (3.6,0.4);

\node[below] at (2,0) {4 units};
\node[right] at (4,1.5) {3 units};
\node[left] at (1.3,1.5) {5 units};
\node[align=center] at (2,-1.2) {\textbf{3-4-5 Triangle} \\ (Right angle construction)};

% Rope depiction
\foreach \i in {0,...,12} {
  \filldraw[black] ({\i*0.33}, -2.2) circle (1pt);
}
\draw[very thick, black] (0,-2.2) -- (4,-2.2);
\node[below] at (2,-2.5) {Rope with 12 equal segments};

% --- Similar Triangles (Right) ---
% Small stick and shadow
\draw[thick] (7,0) -- (7,1.5); % stick
\draw[thick] (7,0) -- (9,0); % shadow
\draw[dashed] (7,1.5) -- (9,0); % hypotenuse
\filldraw[black] (7,1.5) circle (1pt);
\node[left] at (7,0.75) {1.5 m stick};
\node[below] at (8,0) {1 m shadow};

% Pyramid and its shadow
\draw[fill=gray!20] (11,0) -- (12.5,4.5) -- (14,0) -- cycle; % pyramid
\draw[thick] (12.5,0) -- (15.5,0); % shadow
\draw[dashed] (12.5,4.5) -- (15.5,0); % hypotenuse
\node[below] at (14,0) {3 m shadow};
\node[right] at (9.5,2.5) {Height = ?};

% Labels
\node[align=center] at (11.5,-1.2) {\textbf{Similar Triangles} \\ (Using shadows to estimate height)};

\end{tikzpicture}
}
\caption{Left: The \textbf{3-4-5 triangle} used by the \textit{harpedonaptae} to construct right angles. \\
Right: \textbf{Shadow ratio trick}—if a 1.5-meter stick casts a 1-meter shadow, and a pyramid casts a 3-meter shadow, then by similar triangles, the pyramid is 4.5 meters tall.}
\end{figure}


They used the principle of similar triangles to measure inaccessible lengths. Want to figure out the width of the Nile without swimming across and possibly being eaten by a crocodile? Simple: plant a stick in the ground, measure its shadow, then compare it to the shadow of something else (like a pyramid, a tree, or your boss). The ratio of shadow to height stays the same --- assuming the sun isn’t gaslighting you that day.

Need to mark out a right angle? Tie a rope into 12 equal segments, form a triangle with sides 3, 4, and 5, and bam --- congrats, you’ve just built a right triangle. You might not get a medal, but you can now draw a rectangular field without anyone accusing you of witchcraft. 

In essence, these rope-stretching wizards turned geometry into a practical survival skill. Where modern surveyors use high-tech tools and CAD software, the harpedonaptae used proportional reasoning to tame the chaos of nature --- and occasionally, the Nile.



At this point, math was still firmly rooted in the real world. And it could have stayed that way.  

But then the Greeks got involved.  

\begin{figure}[H]
\centering
\begin{tikzpicture}[every node/.style={font=\footnotesize}]

% Panel 1 — Egyptian surveyor doing land ratios
\comicpanel{0}{4}
  {Surveyor}
  {Farmer}
  {Your new field is 2 units long and 1 unit wide. Congrats, you now own a rectangle.}
  {(0,-0.5)}

% Panel 2 — Engineer using ratios
\comicpanel{6.5}{4}
  {Engineer}
  {Foreman}
  {If the base is twice the height, the ramp won’t collapse. Probably.}
  {(0,-0.5)}

% Panel 3 — Egyptian priest estimating golden ratio
\comicpanel{0}{0}
  {Priest}
  {Apprentice}
  {If we keep this ratio, the pyramid looks divine. Or at least intimidating.}
  {(0,0.8)}

% Panel 4 — Greek philosopher enters
\comicpanel{6.5}{0}
  {Greek}
  {Everyone}
  {What if numbers aren’t just useful, but… beautiful?}
  {(0,0.8)}

\end{tikzpicture}
\caption{Egyptians used ratios to measure fields. Greeks used them to measure the soul.}
\end{figure}


\begin{tcolorbox}[colback=blue!5!white, colframe=blue!50!black, title={Historical Sidebar: How the Greeks Measured the Earth — For All the Wrong Reasons}]

  You’ve probably heard the fun fact that the ancient Greeks knew the Earth was round — and even managed to calculate its circumference. But you might have wondered: \textit{How could they possibly know that without satellites, planes, or Google Earth?}

  \medskip
  
  The answer is a mix of brilliant reasoning, clever use of ratios, and a cosmology that was more philosophy than physics.

  \medskip
  
  For the Greeks, especially thinkers like \textbf{Pythagoras} and \textbf{Plato}, the Earth wasn’t round because of empirical evidence. It was round because the sphere was the most \textbf{perfect} shape. In their worldview, the cosmos was built from ideal forms — and a flat, jagged Earth simply didn’t fit their aesthetic. The gods, after all, wouldn’t settle for bad geometry.

  \medskip
  
  Then came \textbf{Eratosthenes}, a librarian with a knack for turning philosophical assumptions into measurable results. He heard that at noon in Syene (modern-day Aswan), the sun shone directly overhead, casting no shadow. But in Alexandria, shadows appeared at the same time of day. 

  \medskip
  
  Rather than shrugging this off as a celestial quirk, Eratosthenes saw an opportunity. Assuming the Earth was a sphere — because of course it was — he measured the angle of the shadow in Alexandria (about 7.2°) and knew the distance between the two cities. Using simple \textbf{ratios}, he scaled that angle up to a full 360° circle and calculated the Earth's circumference with surprising accuracy.

  \medskip
  
  In short, \textbf{they got the right answer... but for the wrong reasons}. The Greeks didn’t prove the Earth was round through modern scientific methods; they started with the assumption that nature loved perfect shapes, and then applied proportional reasoning to make the math work.

\end{tcolorbox}
