\subsection{Applications at the Time}

KL divergence was not born from abstract curiosity. It emerged from the crucible of \textbf{military cryptography} and \textbf{statistical surveillance}—domains where the difference between truth and assumption wasn’t philosophical, but operational. Whether intercepting Soviet ciphers or scanning signals for anomalies, analysts needed a way to quantify just how dangerous it was to be wrong. KL divergence gave them that tool: a way to compare models to reality with cold, statistical rigor.

And then came McCarthyism. The paranoia that once fueled cryptanalysis turned inward, scanning not enemy codes but American citizens for ideological impurity. Surveillance expanded. Inference became suspicion. The same mathematical tools that once helped defend freedom were now repurposed to erode it.

\medskip

\noindent\textbf{The moral?} Once you unleash a technical idea—whether it’s a theorem, a model, or a divergence—you don’t control how it gets used. Kullback and Leibler’s work made error quantifiable. It also made it weaponizable.

\begin{quote}
In the end, KL divergence wasn’t just a tool for measuring information. It became a reminder that the line between knowledge and power is thin—  
and sometimes, drawn in logarithms.
\end{quote}





\begin{figure}[H]
\centering

% === First row ===
\begin{subfigure}[t]{0.45\textwidth}
\centering
\begin{tikzpicture}
  \comicpanel{0}{0}
    {Statistician}
    {}
    {\footnotesize We built KL divergence to quantify how wrong your beliefs are.}
    {(0,-0.6)}
\end{tikzpicture}
\caption*{Born in math, raised in war.}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.45\textwidth}
\centering
\begin{tikzpicture}
  \comicpanel{0}{0}
    {NSA Analyst}
    {}
    {\footnotesize It’s perfect. Now we can measure ideological deviation. In real time.}
    {(0,-0.6)}
\end{tikzpicture}
\caption*{From enemy codes to citizen thoughts.}
\end{subfigure}

\vspace{1em}

% === Second row ===
\begin{subfigure}[t]{0.45\textwidth}
\centering
\begin{tikzpicture}
  \comicpanel{0}{0}
    {Congressman}
    {}
    {\footnotesize If their beliefs diverge from the patriotic line, we log-transform them into suspects.}
    {(0,-0.6)}
\end{tikzpicture}
\caption*{Logarithms make great surveillance filters.}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.45\textwidth}
\centering
\begin{tikzpicture}
  \comicpanel{0}{0}
    {Kullback}
    {}
    {\footnotesize I meant this to find the truth. But truth seems to have... multiple interpretations.}
    {(0,-0.6)}
\end{tikzpicture}
\caption*{Haunted by entropy.}
\end{subfigure}

\caption{KL divergence: First used to measure signal anomalies. Then to flag ideological ones.}
\end{figure}


