\begin{thebibliography}{}

    %%
    %% A
    %%

    \bibitem{aldridge2013}
    Aldridge, I. (2013). \textit{High-Frequency Trading: A Practical Guide to Algorithmic Strategies and Trading Systems} (2nd ed.). Wiley.

    \bibitem{allen2005}
    Allen, E. S., Atkins, D. C., Baucom, D. H., Snyder, D. K., Gordon, K. C., \& Glass, S. P. (2005). Intrapersonal, interpersonal, and contextual factors in engaging in and responding to extramarital involvement. \textit{Clinical Psychology: Science and Practice, 12}(2), 101–130.

    \bibitem{aron1997}
    Aron, A., Norman, C. C., Aron, E. N., McKenna, C., \& Heyman, R. E. (1997). Couple's shared participation in novel and arousing activities and experienced relationship quality. \textit{Journal of Personality and Social Psychology, 78}(2), 273--284. https://doi.org/10.1037/0022-3514.78.2.273
    
    \bibitem{atkins2001}
    Atkins, D. C., Baucom, D. H., \& Jacobson, N. S. (2001). Understanding infidelity: Correlates in a national random sample. \textit{Journal of Family Psychology, 15}(4), 735–749.

    \bibitem{axelrod1984}
    Axelrod, R. (1984). \textit{The evolution of cooperation}. Basic Books.

    %%
    %% B
    %%

    \bibitem{bain2020}
    Bain \& Company. (2020). \textit{How to build a SaaS business that scales}. Retrieved from https://www.bain.com/

    \bibitem{ballard2020}
    Ballard, E. D., \& Zarate, C. A. (2020). The role of dissociation in ketamine’s antidepressant effects. *Nature Communications, 11*(1), 6431. https://doi.org/10.1038/s41467-020-20288-z

    \bibitem{bandura1999}
    Bandura, A. (1999). Moral disengagement in the perpetration of inhumanities. \textit{Personality and Social Psychology Review, 3}(3), 193–209. https://doi.org/10.1207/s15327957pspr0303\_3

    \bibitem{baumeister1991}
    Baumeister, R. F. (1991). \textit{Escaping the self: Alcoholism, spirituality, masochism, and other flights from the burden of selfhood}. Basic Books.
    
    \bibitem{baumeister2004}
    Baumeister, R. F., \& Vohs, K. D. (2004). Sexual economics: Sex as female resource for social exchange in heterosexual interactions. \textit{Personality and Social Psychology Review, 8}(4), 339–363.
    
    \bibitem{bazerman2011}
    Bazerman, M. H., \& Tenbrunsel, A. E. (2011). \textit{Blind spots: Why we fail to do what's right and what to do about it}. Princeton University Press.

    \bibitem{bedi2010}
    Bedi, G., Hyman, D., \& de Wit, H. (2010). Is ecstasy an "empathogen"? Effects of MDMA on prosocial feelings and identification of emotional states in others. *Biological Psychiatry, 68*(12), 1134–1140. https://doi.org/10.1016/j.biopsych.2010.08.003

    \bibitem{benaroya2022}
    Benaroya, D., \& Tan, M. (2022). \textit{SaaS metrics that matter: A playbook for investors and operators}. Sequoia Capital.
    
    \bibitem{berger1966}
    Berger, P. L., \& Luckmann, T. (1966). \textit{The social construction of reality: A treatise in the sociology of knowledge}. Anchor Books.

    \bibitem{bilton2013}
    Bilton, N. (2013, April 28). Disruptions: What happened to Color? No one cared. \textit{The New York Times}. Retrieved from \url{https://www.nytimes.com}

    \bibitem{blank2013}
    Blank, S. (2013). \textit{The Startup Owner's Manual: The Step-by-Step Guide for Building a Great Company}. K\&S Ranch.

    \bibitem{bomfim2005}
    Bomfim, A. N. (2005). \textit{Understanding Credit Derivatives and Related Instruments}. Academic Press.

    \bibitem{bowlby1969}
    Bowlby, J. (1982). \textit{Attachment and loss: Vol. 1. Attachment} (2nd ed.). Basic Books. (Original work published 1969)

    \bibitem{bowlby1988}
    Bowlby, J. (1988). \textit{A secure base: Parent-child attachment and healthy human development}. Basic Books.
    
    \bibitem{bourdieu1984}
    Bourdieu, P. (1984). \textit{Distinction: A social critique of the judgement of taste} (R. Nice, Trans.). Harvard University Press.

    \bibitem{bourne2015}
    Bourne, A., Reid, D., Hickson, F., Torres-Rueda, S., Steinberg, P., \& Weatherburn, P. (2015). “Chemsex” and harm reduction need among gay men in South London. \textit{International Journal of Drug Policy, 26}(12), 1171–1176.
    
    \bibitem{buss1997}
    Buss, D. M., \& Shackelford, T. K. (1997). Susceptibility to infidelity in the first year of marriage. \textit{Journal of Research in Personality, 31}(2), 193–221.

    %%
    %% C 
    %%

    \bibitem{campbell1949}
    Campbell, J. (1949). The hero with a thousand faces. Pantheon Books.   

    \bibitem{chang2018}
    Chang, E. (2018). \textit{Brotopia: Breaking up the boys' club of Silicon Valley}. Portfolio/Penguin.

    \bibitem{choudhry2004}
    Choudhry, M. (2004). \textit{The Credit Default Swap Basis}. Bloomberg Press.

    \bibitem{chu2008}
    Chu, P. S., Ma, W. K., Wong, S. C., Chu, R. W., Cheng, C. H., Wong, S., ... \& Yew, D. T. (2008). The destruction of the lower urinary tract by ketamine abuse: A new syndrome?. \textit{BJU International, 102}(11), 1616–1622.

    \bibitem{christensen1997}
    Christensen, C. M. (1997). \textit{The innovator's dilemma: When new technologies cause great firms to fail}. Harvard Business School Press.

    \bibitem{christensen2020}
    Christensen, J., Shaxson, N., \& Wigan, D. (2020). The Finance Curse: How Global Finance is Making Us All Poorer. \textit{Cornell University Press}.

    \bibitem{cialdini2001}
    Cialdini, R. B. (2001). \textit{Influence: Science and practice} (4th ed.). Allyn \& Bacon.
    
    \bibitem{cialdini2004}
    Cialdini, R. B., \& Goldstein, N. J. (2004). Social influence: Compliance and conformity. \textit{Annual Review of Psychology, 55}, 591–621.

    \bibitem{cooley1902}
    Cooley, C. H. (1902). \textit{Human nature and the social order}. Scribner’s.

    \bibitem{craven2020fashion}
    Craven, J. (2020, May 19). Fashion's love affair with ugliness. \textit{i-D Magazine}. Retrieved from \url{https://i-d.vice.com/en/article/k7ek4m/fashion-ugliness-aesthetic}

    \bibitem{croll2013}
    Croll, A., \& Yoskovitz, B. (2013). \textit{Lean analytics: Use data to build a better startup faster}. O’Reilly Media.

    \bibitem{cutler2003}
    Cutler, B. (2003). \textit{Sexual subcultures}. In A. D. Frank \& S. Wilson (Eds.), \textit{Sexualities: Critical concepts in sociology} (Vol. 3, pp. 145--166). Routledge.
    
    \bibitem{czarniawska1997}
    Czarniawska, B. (1997). \textit{Narrating the organization: Dramas of institutional identity}. University of Chicago Press.

    %%
    %% D 
    %%

    \bibitem{dalenberg2012}
    Dalenberg, C. J., Brand, B. L., Gleaves, D. H., Zanarini, M. C., \& Chu, J. A. (2012). Evaluation of the evidence for the trauma and fantasy models of dissociation. \textit{Psychological Bulletin, 138}(3), 550–588. https://doi.org/10.1037/a0027447

    \bibitem{dames2022}
    Dames, S. A., Wallace, R. T., \& Schweitzer, J. B. (2022). Psychedelic-assisted psychotherapy in women with PTSD: Emerging evidence and therapeutic potential. \textit{Frontiers in Psychology, 13}, 856412.

    \bibitem{dakwar2014}
    Dakwar, E., Anerella, C., Hart, C. L., Levin, F. R., \& Nunes, E. V. (2014). Therapeutic infusions of ketamine: Do the psychoactive effects matter? \textit{Drug and Alcohol Dependence, 136}, 153–157.

    \bibitem{daminger2019}
    Daminger, A. (2019). The cognitive dimension of household labor. \textit{Sex Roles}, 81(7–8), 686–703.

    \bibitem{desai2007}
    Desai, M. A., Dyck, A., \& Zingales, L. (2007). Theft and taxes. \textit{Journal of Financial Economics, 84}(3), 591–623.

    \bibitem{dixit2008}
    Dixit, A. K., \& Nalebuff, B. J. (2008). \textit{The art of strategy: A game theorist’s guide to success in business and life}. W. W. Norton.

    \bibitem{domhoff2014}
    Domhoff, G. W. (2014). \textit{Who rules America? The triumph of the corporate rich}. McGraw-Hill Education.

    %%
    %% E 
    %%

    \bibitem{efrati2008}
    Efrati, A. (2008, June 6). Broadcom co-founder faces drug charges. \textit{The Wall Street Journal}. https://www.wsj.com/articles/SB121267331823324051
    
    \bibitem{ehrlich2023}
    Ehrlich, K. (2023). Invisible labor and gender dynamics in professional and domestic life. \textit{Annual Review of Sociology}, 49, 121–143.

    \bibitem{eisenmann2013}
    Eisenmann, T. (2013). \textit{Platform envelopment}. Harvard Business School Case 807-049.

    \bibitem{escalas2005}
    Escalas, J. E., \& Bettman, J. R. (2005). Self-construal, reference groups, and brand meaning. \textit{Journal of Consumer Research, 32}(3), 378--389. https://doi.org/10.1086/497549

    %%
    %% F 
    %%

    \bibitem{feder2014}
    Feder, A., Parides, M. K., Murrough, J. W., Perez, A. M., Morgan, J. E., Saxena, S., ... \& Charney, D. S. (2014). Efficacy of intravenous ketamine for treatment of chronic posttraumatic stress disorder: A randomized clinical trial. \textit{JAMA Psychiatry, 71}(6), 681–688.

    \bibitem{feld2016}
    Feld, B., \& Mendelson, J. (2016). \textit{Venture Deals: Be Smarter Than Your Lawyer and Venture Capitalist} (3rd ed.). Wiley.
    
    \bibitem{festinger1957}
    Festinger, L. (1957). \textit{A theory of cognitive dissonance}. Stanford University Press.

    \bibitem{fiegerman2017}
    Fiegerman, S. (2017, April 19). Juicero responds to video of people squeezing its juice packs by hand. \textit{Mashable}. Retrieved from \url{https://mashable.com}

    \bibitem{fillmore2003}
    Fillmore, M. T. (2003). Drug abuse as a problem of impaired control: Current approaches and findings. *Behavioral and Cognitive Neuroscience Reviews, 2*(3), 179–197. https://doi.org/10.1177/1534582303257007

    \bibitem{finkel2014}
    Finkel, E. J., Slotter, E. B., Luchies, L. B., Walton, G. M., \& Gross, J. J. (2014). A brief intervention to promote conflict reappraisal preserves marital quality over time. \textit{Psychological Science, 24}(9), 1595–1601. https://doi.org/10.1177/0956797612474938

    \bibitem{frank1999luxury}
    Frank, R. H. (1999). \textit{Luxury fever: Why money fails to satisfy in an era of excess}. Princeton University Press.

    \bibitem{freud1905}
    Freud, S. (1953). Fragment of an analysis of a case of hysteria (J. Strachey, Trans.). In J. Strachey (Ed.), \textit{The standard edition of the complete psychological works of Sigmund Freud} (Vol. 7, pp. 1–122). Hogarth Press. (Original work published 1905)

    \bibitem{freud1926}
    Freud, A. (1959). \textit{The ego and the mechanisms of defence} (C. Baines, Trans.). Hogarth Press. (Original work published 1926)

    \bibitem{fridman2021style}
    Fridman, L. (2021, February 1). \textit{The power of uniforms in the age of Silicon Valley minimalism}. Medium. Retrieved from \url{https://leefridman.medium.com/the-power-of-uniforms-in-silicon-valley-df322adf3ea6}

    \bibitem{friedman2021crocs}
    Friedman, V. (2021, June 30). Crocs are back. And they’re cool. Seriously. \textit{The New York Times}. Retrieved from \url{https://www.nytimes.com/2021/06/30/style/crocs-fashion.html}


    \bibitem{froese2020}
    Froese, T., Iwata, M., Ikegami, T., \& Urushibata, N. (2020). The effect of ketamine on temporal structure of conscious experience: A neurophenomenological study. \textit{Neuroscience of Consciousness, 2020}(1), niaa006.

    %%
    %% G 
    %%
    
    \bibitem{galinsky2006}
    Galinsky, A. D., Magee, J. C., Inesi, M. E., \& Gruenfeld, D. H. (2006). Power and perspectives not taken. \textit{Psychological Science}, 17(12), 1068–1074.

    \bibitem{gallagher2017}
    Gallagher, L. (2017). \textit{The Airbnb story: How three ordinary guys disrupted an industry, made billions... and created plenty of controversy}. Houghton Mifflin Harcourt.
    
    \bibitem{gee2004}
    Gee, J. P. (2004). \textit{Situated language and learning: A critique of traditional schooling}. Routledge.

    \bibitem{gelles2019vest}
    Gelles, D. (2019, April 5). Patagonia is cracking down on how its power vests are used. \textit{The New York Times}. Retrieved from \url{https://www.nytimes.com/2019/04/05/business/patagonia-fleece-vest-finance.html}

    \bibitem{ghosh2019}
    Ghosh, A. (2019). \textit{The SaaS growth playbook}. TechCrunch. Retrieved from https://techcrunch.com/
    
    \bibitem{glass1985}
    Glass, S. P., \& Wright, T. L. (1985). Sex differences in type of extramarital involvement and marital dissatisfaction. \textit{Sex Roles, 12}(9–10), 1101–1120.
    
    \bibitem{glass1992}
    Glass, S. P., \& Wright, T. L. (1992). Justifications for extramarital relationships: The association between attitudes, behaviors, and gender. \textit{Journal of Sex Research, 29}(3), 361–387.
    
    \bibitem{goffman1959}
    Goffman, E. (1959). \textit{The presentation of self in everyday life}. Anchor Books.

    \bibitem{gomber2011}
    Gomber, P., Arndt, B., Lutat, M., \& Uhle, T. (2011). High-frequency trading. \textit{ZfbF}, \textit{63}(9), 237–258. https://doi.org/10.1007/s10203-011-1012-z

    \bibitem{gopnik2010}
    Gopnik, A. (2010). \textit{The table comes first: Family, France, and the meaning of food}. Knopf.

    \bibitem{graeber2011}
    Graeber, D. (2011). \textit{Debt: The first 5{,}000 years}. Melville House.

    \bibitem{graham2020}
    Graham, P. (2020). \textit{Startup = Growth}. Retrieved from https://paulgraham.com/growth.html

    \bibitem{granata2021luxury}
    Granata, F. (2021). \textit{Luxury and irony: The rise of ugly fashion}. London: Bloomsbury Publishing.

    \bibitem{granqvist2004}
    Granqvist, P., \& Kirkpatrick, L. A. (2004). Religious conversion and perceived childhood attachment: A meta-analysis. \textit{International Journal for the Psychology of Religion, 14}(4), 223–250.

    \bibitem{granqvist2006}
    Granqvist, P. (2006). On the relation between secular and divine relationships: An emerging attachment perspective and a critique of the “depth” approaches. \textit{The International Journal for the Psychology of Religion, 16}(1), 1–18.

    \bibitem{guttentag2015}
    Guttentag, D. (2015). Airbnb: Disruptive innovation and the rise of an informal tourism accommodation sector. \textit{Current Issues in Tourism}, 18(12), 1192–1217. https://doi.org/10.1080/13683500.2013.827159

    %%
    %% H 
    %%

    \bibitem{hall1976}
    Hall, E. T. (1976). \textit{Beyond culture}. Anchor Books.

    \bibitem{han2010signaling}
    Han, Y. J., Nunes, J. C., \& Drèze, X. (2010). Signaling status with luxury goods: The role of brand prominence. \textit{Journal of Marketing}, \textit{74}(4), 15--30. https://doi.org/10.1509/jmkg.74.4.15

    \bibitem{harrington2016}
    Harrington, B. (2016). \textit{Capital without borders: Wealth managers and the one percent}. Harvard University Press.

    \bibitem{harvey2005}
    Harvey, D. (2005). \textit{A brief history of neoliberalism}. Oxford University Press.

    \bibitem{hayes2017crocs}
    Hayes, A. (2017, October 1). Balenciaga’s \$850 Crocs are here to haunt your dreams. \textit{The Cut}. Retrieved from \url{https://www.thecut.com/2017/10/balenciaga-platform-crocs-paris-fashion-week.html}

    \bibitem{healy2003}
    Healy, P. M., \& Palepu, K. G. (2003). The fall of Enron. \textit{Journal of Economic Perspectives, 17}(2), 3–26.

    \bibitem{herman1992}
    Herman, J. L. (1992). \textit{Trauma and recovery: The aftermath of violence—from domestic abuse to political terror}. Basic Books.

    \bibitem{hull2012}
    Hull, J. C. (2012). \textit{Options, Futures, and Other Derivatives} (8th ed.). Pearson.

    \bibitem{ho2009}
    Ho, K. (2009). \textit{Liquidated: An ethnography of Wall Street}. Duke University Press.

    \bibitem{hochschild1983}
    Hochschild, A. R. (1983). \textit{The managed heart: Commercialization of human feeling}. University of California Press.

    \bibitem{hoffman2014}
    Hoffman, R., Casnocha, B., \& Yeh, C. (2014). \textit{The Alliance: Managing Talent in the Networked Age}. Harvard Business Review Press.

    \bibitem{hoffman2018}
    Hoffman, R., \& Yeh, C. (2018). \textit{Blitzscaling: The lightning-fast path to building massively valuable companies}. Currency.

    \bibitem{holmes2005}
    Holmes, E. A., Brown, R. J., Mansell, W., Fearon, R. P., Hunter, E. C., Frasquilho, F., \& Oakley, D. A. (2005). Are there two qualitatively distinct forms of dissociation? A review and some clinical implications. \textit{Clinical Psychology Review, 25}(1), 1–23. https://doi.org/10.1016/j.cpr.2004.08.006

    %%
    %% I 
    %%

    %%
    %% J 
    %%

    \bibitem{janoff1992}
    Janoff-Bulman, R. (1992). \textit{Shattered assumptions: Towards a new psychology of trauma}. Free Press.

    \bibitem{jobst2007}
    Jobst, A. A. (2007). Structured finance, credit risk transfer and financial stability. \textit{Financial Markets, Institutions \& Instruments, 16}(1), 1–69. https://doi.org/10.1111/j.1468-0416.2007.00116.x

    \bibitem{jung1968}
    Jung, C. G. (1968). \textit{The archetypes and the collective unconscious} (2nd ed.). Princeton University Press.

    \bibitem{jung1969}
    Jung, C. G. (1969). Symbols of transformation (R. F. C. Hull, Trans.; 2nd ed.). Princeton University Press. (Original work published 1953)

    %%
    %% K 
    %%
    
    \bibitem{kahneman2011}
    Kahneman, D. (2011). \textit{Thinking, fast and slow}. Farrar, Straus and Giroux.

    \bibitem{kang2021}
    Kang, C. (2021, March 5). In Silicon Valley, party culture is thriving in secret. \textit{The New York Times}. https://www.nytimes.com/2021/03/05/technology/silicon-valley-parties-covid.html

    \bibitem{khandani2011}
    Khandani, A. E., \& Lo, A. W. (2011). What happened to the quants in August 2007? Evidence from factors and transactions data. \textit{Journal of Financial Markets}, \textit{14}(1), 1–46. https://doi.org/10.1016/j.finmar.2010.07.005
    
    \bibitem{khan2015}
    Khan, M. (2015). \textit{The Business of Venture Capital: Insights from Leading Practitioners on the Art of Raising a Fund, Deal Structuring, Value Creation, and Exit Strategies}. Wiley.

    \bibitem{khan2011}
    Khan, S. R. (2011). \textit{Privilege: The making of an adolescent elite at St. Paul's School}. Princeton University Press.

    \bibitem{kim2016}
    Kim, T. W. (2016). Acquihires and the market for employees. \textit{Harvard Business Review}. Retrieved from \url{https://hbr.org}

    \bibitem{kirkpatrick1992}
    Kirkpatrick, L. A. (1992). An attachment-theory approach to the psychology of religion. \textit{International Journal for the Psychology of Religion, 2}(1), 3–28.

    \bibitem{kirkpatrick2005}
    Kirkpatrick, L. A. (2005). \textit{Attachment, evolution, and the psychology of religion}. Guilford Press.

    \bibitem{knudson2010}
    Knudson-Martin, C., \& Huenergardt, D. (2010). A socio-emotional approach to couple therapy: Linking social context and couple interaction. \textit{Family Process, 49}(3), 369–386. https://doi.org/10.1111/j.1545-5300.2010.01327.x

    \bibitem{krystal1994}
    Krystal, J. H., Karper, L. P., Seibyl, J. P., Freeman, G. K., Delaney, R., Bremner, J. D., ... \& Charney, D. S. (1994). Subanesthetic effects of the noncompetitive NMDA antagonist, ketamine, in humans. \textit{Archives of General Psychiatry, 51}(3), 199–214.

    \bibitem{krystal2019}
    Krystal, J. H., Abdallah, C. G., Sanacora, G., Charney, D. S., \& Duman, R. S. (2019). Ketamine: A paradigm shift for depression research and treatment. \textit{Neuron, 101}(5), 774–778.

    %%
    %% L 
    %%

    \bibitem{langdridge2007}
    Langdridge, D., \& Barker, M. (Eds.). (2007). \textit{Safe, sane and consensual: Contemporary perspectives on sadomasochism}. Palgrave Macmillan.

    \bibitem{lanius2011}
    Lanius, R. A., Bluhm, R. L., \& Frewen, P. A. (2011). How understanding the neurobiology of complex post-traumatic stress disorder can inform clinical practice: A social cognitive and affective neuroscience approach. \textit{Acta Psychiatrica Scandinavica, 124}(5), 331–348. https://doi.org/10.1111/j.1600-0447.2011.01755.x

    \bibitem{lammers2010}
    Lammers, J., Stoker, J. I., \& Stapel, D. A. (2010). Power and behavioral approach orientation in sexual harassment. \textit{Journal of Personality and Social Psychology, 99}(4), 601–610.

    \bibitem{lasch1979}
    Lasch, C. (1979). \textit{The culture of narcissism: American life in an age of diminishing expectations}. W. W. Norton.
    
    \bibitem{lave1991}
    Lave, J., \& Wenger, E. (1991). \textit{Situated learning: Legitimate peripheral participation}. Cambridge University Press.

    \bibitem{levi2006}
    Levi, M., \& Reuter, P. (2006). Money Laundering. \textit{Crime and Justice, 34}, 289–375.

    \bibitem{liebrenz2009}
    Liebrenz, M., Stohler, R., \& Borgeat, A. (2009). Recurrent nightmares and other psychiatric symptoms after administration of a subanesthetic dose of ketamine. \textit{World Journal of Biological Psychiatry, 10}(4), 847–849.

    \bibitem{lefcourt2017}
    Lefcourt, M. (2017). Lessons from Stripe: Focus your product on a real user pain. \textit{TechCrunch}. Retrieved from \url{https://techcrunch.com}

    \bibitem{lieberman1988a}
    Lieberman, M. B., \& Montgomery, D. B. (1988). First-mover advantages. \textit{Strategic Management Journal}, 9(S1), 41–58. https://doi.org/10.1002/smj.4250090706

    \bibitem{lieberman1998b}
    Lieberman, M. B., \& Montgomery, D. B. (1998). First-mover (dis)advantages: Retrospective and link with the resource-based view. \textit{Strategic Management Journal}, 19(12), 1111–1125. https://doi.org/10.1002/(SICI)1097-0266(1998120)19:12<1111::AID-SMJ21>3.0.CO;2-W

    \bibitem{london2004}
    London, E. D., Simon, S. L., Berman, S. M., Mandelkern, M. A., Lichtman, A. M., Bramen, J., \& Rawson, R. A. (2004). Mood disturbances and regional cerebral metabolic abnormalities in recently abstinent methamphetamine abusers. *Archives of General Psychiatry, 61*(1), 73–84. https://doi.org/10.1001/archpsyc.61.1.73

    \bibitem{lopez2018}
    Lopez, G. (2018, February 13). Silicon Valley’s elite parties aren’t just parties. They’re how power works. \textit{Vox}. https://www.vox.com/technology/2018/2/13/17009070/brotopia-emily-chang-book-silicon-valley-sexism

    \bibitem{lounsbury2001}
    Lounsbury, M., \& Glynn, M. A. (2001). Cultural entrepreneurship: Stories, legitimacy, and the acquisition of resources. \textit{Strategic Management Journal, 22}(6‐7), 545–564.

    \bibitem{luft1955}
    Luft, J., \& Ingham, H. (1955). The Johari Window: A graphic model of awareness in interpersonal relations. In L. Luft (Ed.), \textit{Proceedings of the western training laboratory in group development}. UCLA.

    %%
    %% M 
    %%

    \bibitem{mackenzie2018}
    MacKenzie, D. (2018). \textit{Material Markets: How Economic Agents Are Constructed}. Oxford University Press.

    \bibitem{mahajan2010}
    Mahajan, S. (2010). \textit{Street-Fighting Mathematics: The Art of Educated Guessing and Opportunistic Problem Solving}. MIT Press.

    \bibitem{malhotra1997}
    Malhotra, A. K., Pinals, D. A., Weingartner, H., Sirocco, K., Missar, C. D., Pickar, D., \& Breier, A. (1997). NMDA receptor function and human cognition: The effects of ketamine in healthy volunteers. \textit{Neuropsychopharmacology, 17}(3), 141–150.

    \bibitem{mark2011}
    Mark, K. P., Janssen, E., \& Milhausen, R. R. (2011). Infidelity in heterosexual couples: Demographic, interpersonal, and personality-related predictors of extradyadic sex. \textit{Archives of Sexual Behavior, 40}(5), 971–982.

    \bibitem{marwick2013}
    Marwick, A. E. (2013). \textit{Status update: Celebrity, publicity, and branding in the social media age}. Yale University Press.

    \bibitem{mauboussin2020}
    Mauboussin, M. J., \& Callahan, D. (2020). \textit{Public to Private Equity in the United States: A Long-Term Look}. Morgan Stanley Research.

    \bibitem{mclean2003}
    McLean, B., \& Elkind, P. (2003). \textit{The smartest guys in the room: The amazing rise and scandalous fall of Enron}. Portfolio.

    \bibitem{mcmillan2012}
    McMillan, R. (2012, May 15). Inside the wild life of tech billionaire Henry Nicholas. \textit{Wired}. https://www.wired.com/2012/05/henry-nicholas-broadcom/

    \bibitem{menkveld2013}
    Menkveld, A. J. (2013). High frequency trading and the new market makers. \textit{Journal of Financial Markets}, \textit{16}(4), 712–740. https://doi.org/10.1016/j.finmar.2013.06.006

    \bibitem{mikulincer2007}
    Mikulincer, M., \& Shaver, P. R. (2007). \textit{Attachment in adulthood: Structure, dynamics, and change}. Guilford Press.

    \bibitem{milgram1974}
    Milgram, S. (1974). \textit{Obedience to authority: An experimental view}. Harper \& Row.

    \bibitem{moazed2016}
    Moazed, A., \& Johnson, N. L. (2016). \textit{Modern monopolies: What it takes to dominate the 21st-century economy}. St. Martin’s Press.

    \bibitem{morgan2004}
    Morgan, C. J., Mofeez, A., Brandner, B., Bromley, L., \& Curran, H. V. (2004). Acute effects of ketamine on memory systems and psychotic symptoms in healthy volunteers. \textit{Neuropsychopharmacology, 29}(1), 208–218.

    \bibitem{morris2019fleece}
    Morris, D. Z. (2019, April 7). The Patagonia fleece vest has become the unofficial uniform of Wall Street. \textit{Fortune}. Retrieved from \url{https://fortune.com/2019/04/07/patagonia-vest-wall-street-uniform/}

    \bibitem{muetzelfeldt2008}
    Muetzelfeldt, L., Kamboj, S. K., Rees, H., Taylor, J., \& Morgan, C. J. A. (2008). Journey through the K-hole: phenomenological aspects of ketamine use. \textit{Drug and Alcohol Dependence, 95}(3), 219–229.

    \bibitem{mullins2009}
    Mullins, J., \& Komisar, R. (2009). \textit{Getting to plan B: Breaking through to a better business model}. Harvard Business Press.

    \bibitem{munroe2015}
    Munroe, R. (2015). \textit{Thing Explainer: Complicated Stuff in Simple Words}. Houghton Mifflin Harcourt.

    %%
    %% N 
    %%

    \bibitem{narang2014}
    Narang, R. K. (2014). \textit{Inside the Black Box: A Simple Guide to Quantitative and High Frequency Trading} (2nd ed.). Wiley.

    \bibitem{nasar1998}
    Nasar, S. (1998). \textit{A beautiful mind}. Simon \& Schuster.

    \bibitem{nash1950}
    Nash, J. (1950). Equilibrium points in n-person games. \textit{Proceedings of the National Academy of Sciences, 36}(1), 48–49. https://doi.org/10.1073/pnas.36.1.48

    \bibitem{neumann1954}
    Neumann, E. (1954). The origins and history of consciousness (R. F. C. Hull, Trans.). Princeton University Press.

    \bibitem{newmahr2011}
    Newmahr, S. (2011). \textit{Playing on the edge: Sadomasochism, risk, and intimacy}. Indiana University Press.

    \bibitem{newmahr2011}
    Newmahr, S. (2011). \textit{Playing on the edge: Sadomasochism, risk, and intimacy}. Indiana University Press.

    %%
    %% O 
    %%

    \bibitem{oecd2011}
    OECD. (2011). \textit{Behind the Corporate Veil: Using Corporate Entities for Illicit Purposes}. Organisation for Economic Co-operation and Development.

    \bibitem{osborne1994}
    Osborne, M. J., \& Rubinstein, A. (1994). \textit{A course in game theory}. MIT Press.

    \bibitem{osterwalder2014}
    Osterwalder, A., Pigneur, Y., Bernarda, G., Smith, A., \& Papadakos, T. (2014). \textit{Value Proposition Design: How to Create Products and Services Customers Want}. Wiley.

    %%
    %% P 
    %%

    \bibitem{partnoy2007}
    Partnoy, F., \& Skeel, D. A. (2007). The promise and perils of credit derivatives. \textit{University of Cincinnati Law Review, 75}(3), 1019–1051.

    \bibitem{picciotto2007}
    Picciotto, S. (2007). \textit{Regulatory networks and global governance}. Transnational Corporations, 16(1), 1–37.

    \bibitem{pomarol2006}
    Pomarol-Clotet, E., Honey, G. D., Murray, G. K., Corlett, P. R., Absalom, A. R., Lee, M., ... \& Fletcher, P. C. (2006). Psychological effects of ketamine in healthy volunteers. \textit{The British Journal of Psychiatry, 189}(2), 173–179.

    \bibitem{porter1980}
    Porter, M. E. (1980). \textit{Competitive Strategy: Techniques for Analyzing Industries and Competitors}. Free Press.

    \bibitem{poundstone2003}
    Poundstone, W. (2003). \textit{How Would You Move Mount Fuji? Microsoft's Cult of the Puzzle -- How the World's Smartest Companies Select the Most Creative Thinkers}. Little, Brown.

    \bibitem{putnam1997}
    Putnam, F. W. (1997). \textit{Dissociation in children and adolescents: A developmental perspective}. Guilford Press.

    %%
    %% Q 
    %%

    \bibitem{quart2018}
    Quart, A. (2018). \textit{Squeezed: Why our families can't afford America}. Ecco.

    %%
    %% R 
    %%

    \bibitem{rao1994}
    Rao, H. (1994). The social construction of reputation: Certification contests, legitimation, and the survival of organizations in the American automobile industry: 1895–1912. \textit{Strategic Management Journal, 15}(S1), 29–44.

    \bibitem{rao2011}
    Rao, R. (2011). \textit{The lean startup funnel: Acquisition, activation, retention, referral, revenue}. Retrieved from https://venturehacks.com

    \bibitem{rensin2015}
    Rensin, D. (2015). \textit{The Uncensored Truth About the Lifestyle}. Medium. https://medium.com

    \bibitem{rhodes1986}
    Rhodes, R. (1986). \textit{The Making of the Atomic Bomb}. Simon \& Schuster.

    \bibitem{ries2011}
    Ries, E. (2011). \textit{The Lean Startup: How Today’s Entrepreneurs Use Continuous Innovation to Create Radically Successful Businesses}. Crown Publishing.
    
    \bibitem{rogoff1990}
    Rogoff, B. (1990). \textit{Apprenticeship in thinking: Cognitive development in social context}. Oxford University Press.

    \bibitem{rogers1951}
    Rogers, C. R. (1951). \textit{Client-centered therapy: Its current practice, implications, and theory}. Houghton Mifflin.

    \bibitem{rothman2001}
    Rothman, R. B., Blough, B. E., \& Baumann, M. H. (2001). Dual dopamine/serotonin releasers: Potential treatment agents for stimulant addiction. *Trends in Pharmacological Sciences, 22*(4), 151–153. https://doi.org/10.1016/S0165-6147(00)01608-0

    %%
    %% S 
    %%

    \bibitem{santos2020}
    Santos, G. M., Do, T. N., Beck, J., Makofane, K., Arreola, S., Pyun, T., \& Ayala, G. (2020). Ketamine use and sexual behavior among men who have sex with men: A global cross-sectional study. \textit{Substance Use \& Misuse, 55}(3), 438–447.

    \bibitem{sayer2005}
    Sayer, A. (2005). \textit{The moral significance of class}. Cambridge University Press.

    \bibitem{schmahl2006}
    Schmahl, C., Greffrath, W., Baumgärtner, U., Schlereth, T., Magerl, W., Philipsen, A., ... \& Bohus, M. (2006). Differential nociceptive deficits in patients with borderline personality disorder and self-injurious behavior: Laser-evoked potentials, spatial discrimination of noxious stimuli, and pain ratings. \textit{Pain, 120}(1), 231–239. https://doi.org/10.1016/j.pain.2005.11.027

    \bibitem{scott2009}
    Scott, J. C. (2009). \textit{The art of not being governed: An anarchist history of upland Southeast Asia}. Yale University Press.
    
    \bibitem{sec2008}
    U.S. Securities and Exchange Commission. (2008). \textit{SEC charges Broadcom Corp. and former executives with options backdating}. https://www.sec.gov/news/press/2008/2008-130.htm

    \bibitem{semple2004}
    Semple, S. J., Patterson, T. L., \& Grant, I. (2004). The context of sexual risk behavior among heterosexual methamphetamine users. *Addictive Behaviors, 29*(4), 807–810. https://doi.org/10.1016/j.addbeh.2004.02.001

    \bibitem{sennett1977}
    Sennett, R. (1977). \textit{The fall of public man}. Knopf.

    \bibitem{sedgwick2003}
    Sedgwick, E. K. (2003). \textit{Touching feeling: Affect, pedagogy, performativity}. Duke University Press.

    \bibitem{sharman2010}
    Sharman, J. C. (2010). Shopping for Anonymous Shell Companies: An Audit Study of Anonymity and Crime in the International Financial System. \textit{Journal of Economic Perspectives, 24}(4), 127–140.

    \bibitem{shahani2007}
    Shahani, R., Streutker, C., Dickson, B., \& Stewart, R. J. (2007). Ketamine-associated ulcerative cystitis: A new clinical entity. \textit{Urology, 69}(5), 810–812.

    \bibitem{shiller2017}
    Shiller, R. J. (2017). \textit{Narrative economics}. American Economic Association Presidential Address.

    \bibitem{siegel1999}
    Siegel, D. J. (1999). \textit{The developing mind: How relationships and the brain interact to shape who we are}. Guilford Press.

    \bibitem{slim1969}
    Slim, I. (2011). \textit{Pimp: The story of my life}. Canongate Books. (Original work published 1969)

    \bibitem{soper2020}
    Soper, S., \& Chapman, L. (2020, December 10). Airbnb’s shares soar in biggest U.S. IPO debut of 2020. \textit{Bloomberg}. Retrieved from https://www.bloomberg.com/

    \bibitem{sorkin2008}
    Sorkin, A. R. (2008, June 6). The curious case of Henry Nicholas. \textit{The New York Times DealBook}. https://dealbook.nytimes.com/2008/06/06/the-curious-case-of-henry-nicholas/

    \bibitem{snorrason2012}
    Snorrason, I., Belleau, E. L., \& Woods, D. W. (2012). Clinical characteristics of hair-pulling disorder and excoriation (skin-picking) disorder. \textit{Child and Adolescent Psychiatric Clinics of North America, 21}(1), 111–133. https://doi.org/10.1016/j.chc.2011.09.004

    \bibitem{srnicek2017}
    Srnicek, N. (2017). \textit{Platform Capitalism}. Polity Press.

    \bibitem{stewart2011pappy}
    Stewart, J. B. (2011, December 1). Pappy Van Winkle, the bourbon everyone wants but no one can get. \textit{The New York Times}. Retrieved from https://www.nytimes.com/2011/12/03/business/pappy-van-winkle-bourbon-in-short-supply.html

    \bibitem{spence1973}
    Spence, M. (1973). Job market signaling. \textit{Quarterly Journal of Economics, 87}(3), 355–374.

    \bibitem{steele1990}
    Steele, C. M., \& Josephs, R. A. (1990). Alcohol myopia: Its prized and dangerous effects. *American Psychologist, 45*(8), 921–933. https://doi.org/10.1037/0003-066X.45.8.921

    %%
    %% T 
    %%
    
    \bibitem{tajfel1986}
    Tajfel, H., \& Turner, J. C. (1986). The social identity theory of intergroup behavior. In S. Worchel \& W. G. Austin (Eds.), \textit{Psychology of intergroup relations} (pp. 7–24). Nelson-Hall.

    \bibitem{taylor2001}
    Taylor, G. W., \& Ussher, J. M. (2001). Making sense of S\&M: A discourse analytic account. \textit{Sexualities, 4}(3), 293--314. \url{https://doi.org/10.1177/136346001004003002}

    \bibitem{taylor2015dress}
    Taylor, K. (2015, July 30). The new power uniform in Silicon Valley is a Patagonia vest. \textit{Business Insider}. Retrieved from \url{https://www.businessinsider.com/silicon-valley-dress-code-patagonia-vest-2015-7}

    \bibitem{toffler2003}
    Toffler, B. L., \& Reingold, J. (2003). \textit{Final accounting: Ambition, greed and the fall of Arthur Andersen}. Broadway Books.

    \bibitem{trivers2011}
    Trivers, R. (2011). \textit{The folly of fools: The logic of deceit and self-deception in human life}. Basic Books.

    \bibitem{tschorn2021idiocracy}
    Tschorn, A. (2021, July 27). How Crocs went from \textit{Idiocracy} joke to fashion runway must-have. \textit{Los Angeles Times}. Retrieved from \url{https://www.latimes.com/lifestyle/story/2021-07-27/how-crocs-went-from-idiocracy-joke-to-fashion-runway-must-have}

    \bibitem{tuckman2011}
    Tuckman, B., \& Serrat, A. (2011). \textit{Fixed Income Securities: Tools for Today's Markets} (3rd ed.). Wiley.

    \bibitem{turner2009}
    Turner, F. (2009). Burning Man at Google: A cultural infrastructure for new media production. \textit{New Media \& Society, 11}(1-2), 73–94. https://doi.org/10.1177/1461444808099575

    \bibitem{tzuo2018}
    Tzuo, T., \& Weisert, G. (2018). \textit{Subscribed: Why the subscription model will be your company's future—and what to do about it}. Portfolio.

    %%
    %% U 
    %%

    \bibitem{ussenate2002}
    U.S. Senate. (2002). \textit{Financial collapse of Enron: Hearing before the Committee on Governmental Affairs}. U.S. Government Printing Office.

    \bibitem{usdistrict2008}
    U.S. District Court. (2008). \textit{United States of America v. Henry T. Nicholas III}, Criminal Complaint No. SACR-08-00139. United States District Court, Central District of California.

    %%
    %% V 
    %%

    \bibitem{van2014}
    van der Kolk, B. A. (2014). \textit{The body keeps the score: Brain, mind, and body in the healing of trauma}. Viking.

    \bibitem{vygotsky1978}
    Vygotsky, L. S. (1978). \textit{Mind in society: The development of higher psychological processes} (M. Cole, V. John-Steiner, S. Scribner, \& E. Souberman, Eds.). Harvard University Press.

    \bibitem{volkow2001}
    Volkow, N. D., Chang, L., Wang, G.-J., Fowler, J. S., Franceschi, D., Sedler, M., ... \& Logan, J. (2001). Loss of dopamine transporters in methamphetamine abusers recovers with protracted abstinence. *Journal of Neuroscience, 21*(23), 9414–9418. https://doi.org/10.1523/JNEUROSCI.21-23-09414.2001

    %%
    %% W 
    %%
    
    \bibitem{watkins2002}
    Watkins, S. (2002). \textit{Testimony before the U.S. House Subcommittee on Oversight and Investigations}. U.S. Government Printing Office.

    \bibitem{wedel2009}
    Wedel, J. R. (2009). \textit{Shadow elite: How the world's new power brokers undermine democracy, government, and the free market}. Basic Books.

    \bibitem{weinstein2008}
    Weinstein, L., \& Adam, J. A. (2008). \textit{Guesstimation: Solving the World's Problems on the Back of a Cocktail Napkin}. Princeton University Press.

    \bibitem{weiss2011}
    Weiss, M. E. (2011). \textit{Techniques of pleasure: BDSM and the circuits of sexuality}. Duke University Press.

    \bibitem{wells2014pappy}
    Wells, P. (2014, November 25). A taste of Pappy Van Winkle, the bourbon that vanishes. \textit{The New Yorker}. Retrieved from https://www.newyorker.com/culture/culture-desk/a-taste-of-pappy-van-winkle-the-bourbon-that-vanishes

    \bibitem{whisman2007}
    Whisman, M. A., Gordon, K. C., \& Chatav, Y. (2007). Predicting sexual infidelity in a population-based sample of married individuals. \textit{Journal of Family Psychology, 21}(2), 320–324.

    \bibitem{wilkinson2017}
    Wilkinson, S. T., Sanacora, G., \& Bloch, M. H. (2017). Ketamine for depression: Safety and efficacy in clinical trials. \textit{Pharmacology \& Therapeutics, 176}, 16–24.

    \bibitem{wolff2017}
    Wolff, M. (2017, August). The world according to Silicon Valley’s secret parties. \textit{Vanity Fair}. https://www.vanityfair.com/news/2017/08/silicon-valley-secret-parties-elites
    
    \bibitem{williams2008}
    Williams, J. C. (2008). \textit{Unbending gender: Why family and work conflict and what to do about it}. Oxford University Press.

    %%
    %% X 
    %%

    %%
    %% Y 
    %%

    \bibitem{yin2017}
    Yin, M. (2017). Engineering the acquihire: How Silicon Valley reconfigures labor. \textit{TripleC: Communication, Capitalism \& Critique, 15}(1), 157–172.

    %%
    %% Z 
    %%

    \bibitem{zarate2006}
    Zarate, C. A., Singh, J. B., Carlson, P. J., Brutsche, N. E., Ameli, R., Luckenbaugh, D. A., ... \& Manji, H. K. (2006). A randomized trial of an N-methyl-D-aspartate antagonist in treatment-resistant major depression. \textit{Archives of General Psychiatry, 63}(8), 856–864.

    \bibitem{zajonc1968}
    Zajonc, R. B. (1968). Attitudinal effects of mere exposure. \textit{Journal of Personality and Social Psychology, 9}(2, Pt.2), 1--27. https://doi.org/10.1037/h0025848

    \bibitem{zider1998}
    Zider, B. (1998). \textit{How venture capital works}. Harvard Business Review, 76(6), 131–139.

    \bibitem{zimbardo2007}
    Zimbardo, P. G. (2007). \textit{The Lucifer effect: Understanding how good people turn evil}. Random House.

    \bibitem{zuckerman1999}
    Zuckerman, E. W. (1999). The categorical imperative: Securities analysts and the illegitimacy discount. \textit{American Journal of Sociology, 104}(5), 1398–1438.

    \bibitem{zucman2015}
    Zucman, G. (2015). \textit{The hidden wealth of nations: The scourge of tax havens}. University of Chicago Press.

    
    
    
    
    
    
    
    
    
    
     
    
     
    

    
    
    
    
        
    
    
    
    

    
    
    
    
    
        

    
    
    
    
     
    
    
    
    
    
        
    
    
    
    
    
    
    
    
    
     
    
    
     


\end{thebibliography}