\subsection{The Logistics Team}

The country club pool shimmered beneath the late afternoon sun, its surface dappled with gold where the 
water caught the light. The haze of summer softened everything — the edges of the lounge chairs, the 
rhythm of the tennis match drifting from the upper courts, even the shrieks of children leaping into 
the deep end.

``Marco!'' shouted Nora, already halfway underwater.

``Polloooooo,'' Oliver replied from behind a pool float, dragging out the vowels like a game show buzzer.

Before the next round could begin, Serena’s twins — Zoe and Miles — burst from the snack bar clutching 
grape popsicles, and dove into the game without pause.

``Team sonar!'' Miles called, belly-flopping into the shallows.

``No teams!'' Nora insisted, splashing furiously. ``You’re all cheaters anyway.''

``Not true,'' Zoe declared, smug behind mirrored goggles. ``I echolocate legally.''

Emma smiled behind her sunglasses. That argument had started in the car two weeks ago, after Serena 
took them all to the lake for paddleboarding. Oliver had claimed ``sonar powers,'' prompting Nora to 
lecture him on echolocation until everyone dissolved into laughter. Today, the line had become canon.
And Nora was theatrically outraged, but secretly delighted to have it stolen and embellished by someone else.

Now the pool echoed with squeals, accusations, and declarations of who was “it,”  
though no one seemed interested in keeping score.

On the sidelines, Emma and Serena lounged like co-conspirators.

It didn’t feel like babysitting. It felt like summer — shared.

Emma reached for her drink --- a grapefruit spritz Serena had recommended (``Not too sweet, not too serious'') 
--- and watched the kids spiral into another round of inside jokes that weren’t really inside anymore. 
Not for her.

She looked back at the pool, where Oliver was now declaring himself a hedge fund shark and trying to 
``short'' Nora’s cannonball.

Emma laughed again. Not because it was funny --- though it was --- but because she understood it. 
All of it.

This wasn’t just a scene she was watching.
It was one she belonged to.

And maybe, for the first time in years, that felt... safe.

``They’ve adopted each other,'' Serena had joked once. ``We’re just the logistics team.''

Today, Serena was lounging beside her, barefoot and sun-drowsy, a linen wrap falling loosely around her 
shoulders. 
She held her glass like an afterthought, eyes hidden behind oversized sunglasses.

Emma glanced over. ``You ever think they’re the ones pulling us together?''

Serena gave the faintest smile. ``If they are, they’re doing a better job than most boardrooms I’ve sat in.''

Emma swirled the ice in her glass and let her gaze linger on the chaos unfolding in the deep end.

``Remember the koi pond incident?'' she said, a grin tugging at the corner of her mouth.

Serena groaned, then chuckled. ``Zoe swore the fish were trying to communicate with her. Miles offered to decode 
it... for a fee.''

``Oliver still thinks he can speak 'koi','' Emma said. ``He tried it last week in the bathtub. Claimed he got 
stock tips.''

Serena snorted. ``Did he short goldfish futures?''

``Only if Nora let him hedge with snack crackers.''

They both laughed, soft and warm.

Serena leaned back, sighing contentedly. ``That night, they all fell asleep in a heap under the dining table. 
Miles had a sock on his hand like it was a puppet.''

Emma nodded. ``Nora told me later they were ‘hedge fund interns’ and had been up past midnight 
‘chasing liquidity’.''

Serena looked over. ``Where do they get this stuff?''

Emma smiled again. ``From us. From each other. From being just close enough to grown-up conversations 
they’re not supposed to understand — but do anyway.''

Serena was quiet for a moment, watching Zoe help Nora build a floating citadel out of noodles and overturned pool floats.

Then: ``Do you think they’ll remember this?''

Emma’s voice softened. ``I hope so. Even if they don’t remember the details. I hope they remember what it 
felt like. This kind of easy. This kind of... belonging.''

Serena turned to face her, finally lifting her sunglasses to meet her eyes.

``That’s what I wanted for them,'' she said. ``Before I ever knew I wanted it for myself.''

Emma didn’t answer right away. She didn’t need to.

Instead, she raised her glass gently toward Serena.  
No toast. No words.

Just a silent acknowledgment... of the scene... of the season... and of the space they’d all made together.

Serena tapped her glass against it.

``To the logistics team,'' she murmured.

``And to the ones running the show,'' Emma replied, nodding toward the water, where Miles had just declared 
himself CFO of the float kingdom.