
\subsection{The Footnotes of Failure}

David sat across the table, the fluorescent lights above humming with the kind of corporate 
indifference he’d grown used to.

The regulator set the file down slowly. Flipped to the last tab.

“Are these your initials?” he asked, pointing to the bottom-right corner of a commit approval screen.

David leaned forward. Paused. Then nodded.
“Yes.”

The regulator didn’t look triumphant. 
“I spoke with the auditors this morning,” he said. “Not the first line guys — the internal 
forensic crew. The ones who come in after the smoke clears.”

He sat back.

“They weren’t looking to fix the system. They came to write the story. And stories need names.”

He let that land.

“They told me how it started. Not with alerts. Not with alarms. But with footprints.”

He opened the folder again, laying it flat between them.

“Timestamps. Code commits. Deployment notes. Nothing dramatic. Just... sequence. Every action 
left a mark.”

David stayed silent.

“They followed the trail to the model,” the regulator continued. “The one that was supposed to 
catch the risk. But it didn’t.”

He tapped once.

“Wrong signals. Wrong timing. Wrong assumptions for that kind of market.”

David inhaled through his nose. Said nothing.

“Then they asked how the model got out there,” the man said. “That’s where it gets… fragile.”

\begin{itemize}
  \item A launch pushed two weeks early.
  \item A code freeze nobody honored.
  \item A patch that bypassed peer review because \textit{‘we had to move fast.’}
\end{itemize}

“And finally,” the regulator said, almost gently now, “they found the sign-off. The click 
that turned it all real.”

He closed the folder.

“Three letters. Lower right. Yours.”

David looked down.

There was no malice. No panic. Just a moment of quiet resignation.

He knows he has to play the part of the villain. 

\medskip

\begin{HistoricalSidebar}{Auditors vs. Regulators — Two Tribes of Postmortem Power}

  When financial systems fail, two professional species arrive: \textbf{auditors} and \textbf{regulators}. 
  Both investigate. Both ask questions. But their mandates — and temperaments — diverge in subtle, consequential ways.

  \medskip
  
  \textbf{Auditors} are internal or contracted examiners. Their job is to verify compliance with stated policies, 
  reconcile transactions, and ensure that procedures — even flawed ones — were followed. They don’t ask whether a 
  rule made sense. They ask whether it was followed and documented.

  \medskip
  
  In the 2001 Enron collapse, Arthur Andersen’s audit teams had documented procedures — but failed to challenge 
  the legitimacy of off-balance-sheet structures. They checked the math. They missed the meaning.

  \medskip
  
  \textbf{Regulators}, on the other hand, arrive on behalf of public institutions. Their mission is broader: 
  assess systemic risk, uncover governance failures, and assign accountability. While auditors scrutinize evidence, 
  regulators write the narrative. Where auditors measure, regulators interpret.

  \medskip
  
  After the 2008 crisis, agencies like the SEC, CFTC, and Financial Crisis Inquiry Commission sought more than numbers: 
  they sought names. Lehman’s liquidity “death spiral,” AIG’s collateral triggers, and Citi’s CDO masking all became 
  regulatory foci not just because rules were broken, but because stories were buried.

  \medskip
  
  In Aurora’s case, the auditors came first. They brought spreadsheets.  
  The regulators came later. They brought subtext.
  
\end{HistoricalSidebar}

\medskip
