
\subsection{The Belief That Broke First}

They were still in the war room. No one had left. The lights had dimmed slightly, or maybe it just felt that way.

The oil chart had stopped updating. Not because the market was closed — just because no one was watching it anymore.

“I don’t get it,” the Analyst said, voice thin. “Why the credit blowout? Logistics weren’t even the trade.”

“They were the hedge,” the Quant said flatly. “They weren’t supposed to move unless energy volatility stayed elevated.”

David didn’t look up. “It wasn’t about the names. It was about the structure.”

Kessler re-entered, carrying a folder that looked like it had been bent in half by someone’s hand. He dropped it on the table.

“The insurance desk just hit their threshold. They’re unrolling the basis book.”

No one spoke.

“They weren’t even long oil,” the Analyst muttered.

“Doesn’t matter,” David said. “They were long on everyone else being long. That’s how consensus trades work. You’re not betting on the story. You’re betting on the bet.”

The Quant leaned forward, rubbing his eyes. “We were positioned for stagflation. High input costs, fractured supply chains, a world that limped along just badly enough to justify overpricing risk.”

The Analyst nodded slowly. “So when peace showed up—”

“—the whole map broke,” David said. “And the models couldn’t reorient fast enough.”

Kessler looked at them, eyes dark. “You’re telling me this entire drawdown came from a fucking joint statement?”

“No,” the Quant said. “It came from everyone realizing they didn’t know what anything was worth without tension.”

The Analyst whispered, “I was modeling rerouted freight for Q4. It’s worthless now. The port rebookings alone—”

“It’s not just you,” David cut in. “The Korean desk had short-dated options on cross-border trucking volatility. The whole position’s toast.”

Kessler looked around the room like he expected someone to laugh. No one did.

“Jesus,” he said. “We weren’t long oil. We were long friction.”

David leaned back. “And friction was the anchor. When it disappeared, the assumptions started floating.”

The Analyst blinked. “So what do we do now?”

Kessler didn’t answer. He just stared at the glass wall like it might open up and swallow the floor beneath them.

David’s voice was quiet now. “Hope the lawyers find slower clocks than the market.”

The Quant, almost to himself: “If the bridge was never broken… how many trades were detours around nothing?”

